An integrated open population distance sampling approach for modelling age-structured populations

### 2024-01-05 Manuscript_Nilsen_Nater.qmd
Edited/rewrote/commented on discussion

Given the available data, the IDSM was not able to estimate a clear effect of small rodent abundance on ptarmigan recruitment (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I. = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`).

@@

We developed a model that use line transect distance sampling survey data combined with data from marked individuals to estimate population abundance, survival probabilities and recruitment rates in an open population setting. We first used simulated data to examine models ability to recover the underlying parameters when they were known. Then we fitted the model to data from an ongoing field study on willow ptarmigan in Norway. In general, the model did a very good job in recreating the underlying parameters when fitted to simulated data where the true values are known. For the simulated data, a relative wide range of parameter values were allowed, and we did not detect any particular problems or biases with respect to this variation in the underlying parameters. When the input data are unbiased (with respect to the underlying model formulation) we therefore consider the model to be able to extract unbiased and meaningful demographic parameters from distance sampling data.

Open population formulations of the distance sampling model has previously been presented (refs), and applied to various ecological systems (refs). Our model extends those previous applications by formulating the underlying process model as a stage-structured matrix model where the matrix elements are represented by annual survival probabilities and recruitment rates. While such models have been available for a range of other modelling frameworks previously (in particular), this is to our knowledge the first explicit use of such a state model in a distance sampling framework. Such a formulation allow us to both to obtain estimates of the underlying demographic rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$). As an example of such environmental variation, we investigated the effects of small rodent abundance on the recruitment rates. The fact that we were able to successfully recover the covariate effect when the model was applied to simulated data, suggest that in fact the model is able to estimate the effect when there is a clear signal in the data. When applied to real data, we did not detect a clear effect of rodent abundance (indexed by the occurrence of rodents on the transect lines) on recruitment rate. This might be somewhat surprising, given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). We speculate that there are (at least) three potential explenations to this result. First, for our specific case study the data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small. It is well known that the amplitude and regularity of the rodent cycles has been fading in recent decades (refs), and our study area might be no exception. Lack of peak rodent years in the time series to which we fitted the model. In addition, previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); we did not fit any weather covariates to the model so there is a possibility that strong effects of spring conditions in certain years have masked any remaining effects of small rodent abundance. Finally, the data set we used is relative short (15 years), leaving us with somewhat limited statistical power. We therefore do not consider our a particularily strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates. In future applications, a stronger test based on more years of data and data from additional areas is encouraged. In e.g. [@Bowler2020] we used data from the same sampling program but from more areas, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate. Our new model allow us to investigate to which extent similar results emerge when we can link the covariate effect to the actual demographic rate and not only to the population growth rate.

One potential source of bias in our model framework is related to the way that we use the distance sampling data to estimate survival and recruitment rates. In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and thsi create a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. As a test of the phenomena, we also assessed how sensitive the model output was with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**). In the version presented in the results section, we made the assumption that these birds were in fact juveniles. In an alternative scenario, we discarded all birds classified as unknowns, and rerun the model (Appendix). As expected the estimated population density was virtually unaffected by this decision, whereas the effect on the demographic parameters were somewhat more affected. Thus biases in the age ratio will affect the demographic rates, but not so much the estimated density. Compared to previous studies on ptarmigan (see e.g. [@Israelsen2020][@Sandercock2011]), we could have expected a higher survival rate indicating that there is some misclassification in the data with respect to age specification. In future developments of the model, explicit accounting for such misclassification will be considered.

-   Both simulated data tests and real data application have shown that mixing is suboptimal for the average recruitment rate parameter (Mu.R). Future work should perhaps look into optimizing sampling? (For biological interpretation that is not a big problem as the yearly recruitment rate estimates are fine... but have to keep in mind when interpreting) <!-- I leave it to you Erlend to decide where, whether, and how much you want to go into this one -->

In summary, we conclude that the model presented here is useful in many ecological systems where the sampling survey data include information about the age-ratio of the observed individuals. Following the integrated modelling philosophy, it also allow for the integration of auxiliary data. In our application here, we integrated data from marked birds by formulating an explicit (known-fate) mark-recapture model that explicitly supported the estimation of survival probabilities in the state process. The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates. Moreover, the model can be extended with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates. Finally, as the underlying model structure is based on a hierarchical model the framework readily support multi-area extensions, where the covariate-effects are efficiently estimated by time-for-space substitution.

>>>>

Given the available data, the IDSM was not able to estimate a clear effect of small rodent abundance on ptarmigan recruitment (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I. = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`, Supplementary Figure "Rep_betaR.R.png").

@@

We developed an integrated model that jointly analyses line transect distance sampling survey data and data from marked individuals to estimate population abundance, survival probabilities ,and recruitment rates over time. We first used simulated data to examine model's ability to recover the underlying parameters when they were known. We then fitted the model to data from an ongoing field study on willow ptarmigan in Norway to showcase its applicability to real wildlife monitoring data.

Open population formulations of the distance sampling model have previously been presented (refs), and applied to various ecological systems (refs). The model presented here extends those previous applications by formulating the underlying population process model as a stage-structured matrix model [@caswell2000] in which the matrix elements are represented by annual survival probabilities and recruitment rates. While this has been the common approach for a range of other statistical modelling frameworks, including the growing suite of models falling into the category of integrated population models [@schaub2022], the integration of mechanistic population models into distance sampling frameworks is rather new. The resulting modelling framework allow us to make maximum use of distance sampling data in combination with auxiliary information both from the distance sampling survey itself (i.e. information on age, sex, etc. of observed animals) and from other types of monitoring, and enables estimation not only of changes in population density but also of underlying vital rates over time.

In general, the model did a very good job at recreating the underlying parameters when fitted to simulated data where the true values are known. The simulated data sets were based on relatively wide ranges of parameter values, yet we did not find the model to have any particular problems or biases recovering the original parameters used in simulations. When the input data are unbiased (with respect to the underlying model formulation) we therefore consider the model to be able to extract unbiased and meaningful demographic parameters from age-structured distance sampling data.

Real data, however, are likely to be subject to certain biases, and one type of bias that is likely to be common for age-structured distance sampling data arises from failure to (correctly) classify the age class of observed individuals. <!-- Maybe add one illustrative sentence here about how/why this happens using our ptarmigan monitoring setup as an example? --> Observations with incorrectly classified age have the potential to introduce bias in the IDSMs relative estimates of survival and recruitment. This is due to the way it usese the distance sampling data to estimate survival and recruitment rates. In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and this creates a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. To get an idea of the potential effect of this on parameter estimation, we checked the sensitivity of the output of the model fit to real data with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**). In the model version presented in the results section, we made the assumption that these birds were in fact juveniles. <!-- Restate here why this is likely / probably a good guess --> Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (Appendix) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival). Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density. Since the proportion of "unknown sex and age" observations in our ptarmigan case study was rather low, potential biases in estimates resulting from age misclassification are expected to be small. Nonetheless, future developments of the IDSM modelling framework should focus on ways of accounting explicitly for misclassification of age class in the field.

<!-- I suggest we start this next paragraph with a couple of general sentences about the density estimates, even if it is just stating that the density estimates correspond closely to what we know/have estimated with other models previously. Maybe also add one sentence or so about the possible biology behind the recent increase. Then moving on to survival and recruitment -->

Compared to previous studies on ptarmigan (see e.g. [@Israelsen2020; @Sandercock2011]), we could have expected somewhat higher estimates of survival probability. One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, may have changed over time. If survival in more recent years, when telemetry data was collected [and the study of e.g. @@Israelsen2020 was carried out] was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower. On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on ptarmigan (+refs). <!-- If we have any "biological" thoughts about why somewhat high recruitment rates may not be that ridiculous here, then that could be added here -->

In addition to estimating demographic rates from line transect data, the IDSM also allows including relevant environmental effects on the demographic rates themselves, and not just on population growth rate as a whole ($\lambda$).
In the ptarmigan case study we thus attempted to investigate the effect of small rodent abundance (approximated as the proportion of transect lines on which rodents were reported each year) on recruitment rate. <!-- Deleted this because we did NOT include the rodent effect in the simulated data tests --> Unfortunately, we were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (Supplementary Figure "Rep_betaR.R.png"). This may seem somewhat surprising at first glance, given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). We speculate that there are (at least) three potential explanations to this. The first is that our covariate data may not have been well suited for estimating effects on recruitment. The data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small otherwise, making for a covariate with relatively little information content. While this may be partially a consequence of how these data are collected, it is also well known that the amplitude and regularity of the rodent cycles has been fading in recent decades (refs), and our study area might be no exception. Lack of peak rodent years in the time series to which we fitted the model may thus also have contributed to making effect estimation challenging. Second, it is possible that rodent effects were obscured by other, potentially stronger, covariate effects. Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that strong unaccounted for effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance. Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates . Taken together we therefore do not consider this study as a particularly strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates. It is worth noting that future applications could increase statistical power by including either more years of data or capitalizing on space-for-time substitution as the Norwegian ptarmigan monitoring programme spans many more locations beyond Lierne. @Bowler2020, for example, used data from the same sampling program but from more areas using a simpler open population DS model, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate. An extension of our IDSM to include data from multiple areas therefore constitutes a promising approach for investigating to which extent similar results emerge by linking environmental covariates to the actual demographic rates and not only just to the resulting population growth rates.

The new IDSM framework presented here is relevant for many wildlife populations that are surveyed using line transect sampling that includes additional information on age, sex, and/or life stages of the observed individuals. Following the integrated modelling philosophy, the IDSM also allows for the integration of auxiliary data. In our application here, we integrated data from radio-telemetry of marked birds, which explicitly supported the estimation of survival probabilities. The IDSM framework is very flexible, however, and open to the inclusion of additional/other auxiliary data that contains information on demographic rates or population size/density. Moreover, hierarchical nature of the model makes is straight forward to adapt to different species and to include different suites of environmental covariates on the demographic rates. Finally, it constitutes a modelling framework that is well suited for extension to multiple areas and thus able to capitalize on space-for-time substitution to produce large-scale and spatially explicit estimates of population density, demographic rates, and environmental effects from large-scale (participatory) monitoring.
